### 2023-12-18 draft_jss/draft_jss.Rmd
update draft

Reservoir Computing (RC) is a prominent machine learning methodology, proposed by @kim_health_2020 and @maass_real-time_2002,
that has gained significant attention in recent years for its ability to
effectively process information generated by dynamical systems. This
innovative approach leverages the dynamics of a high-dimensional
reservoir to perform complex computations and solve various tasks based
on the response of this dynamical system to input signals. RC has demonstrated its efficacy in tackling various challenges, encompassing pattern classification and time series forecasting in applications ranging from electrocardiogram analysis to bird calls, power plants, internet traffic, stock prices, and beyond [@tanaka_recent_2019].

RC is a kind of recurrent neural networks (RNNs) where the internal states, denoted as the reservoir, is randomly generated and solely the final layer is trained. The reservoir projects temporal input signals onto a high-dimensional feature space facilitating the learning of non-linear and temporal interactions. This offers the advantage of decreasing the computing time while consistently maintaining robust performance compared to conventional RNNs [@shahi_prediction_2022, @vlachas_backpropagation_2020].

The utilization of RC poses challenges on multiple fronts. Firstly, its stochastic nature, attributed to the random aspect of the reservoir, introduces unpredictability into the results, even with well-chosen hyperparameters. Secondly, RC relies on numerous hyperparameters to establish connections between neurons. Guidance on determining these hyperparameters is limited, often requiring users to adopt a wrapper approach where combinations of hyperparameter performance are assessed on a training set before selecting the optimal combination. Thirdly, there is a current absence of implementation in R, rendering the method challenging for users unfamiliar with Python [@trouvain_reservoirpy_2020, @steiner_pyrcn_2022] or Julia [@martinuzzi_reservoircomputingjl_2022].

In this paper, we endeavor to delve into these challenges, offering comprehensive guidance to assist new users in maximizing the benefits of RC. Initially, a broad introduction to reservoir computing is presented, accompanied by a tutorial on its application using `reservoirnet`, an R package built upon the `reservoirPy` Python module [@trouvain_reservoirpy_2020]. Subsequently, we investigate the various challenges associated with the use of RC for forecasting COVID-19 hospitalizations. This exploration includes detailed guidance on the modeling strategy, the selection of hyperparameters, and the implementation process.

@@

RC is a machine learning algorithm composed of an input layer denoted
$u(t)$ which is randomly transformed by the reservoir into an activation
state denoted $x(t)$ which is fed to a ridge penalized linear regression
trained to forecast the outcome denoted as $y(t)$ as depicted at figure
\@ref(fig:rcpresentation).

```{r rcpresentation, echo=FALSE, fig.cap="my caption", out.width = '90%', fig.cap="Reservoir computing is composed of an input layer, a reservoir and an output layer. Connection between input layer and reservoir and inside reservoir are random. Only the output layer is optimized based on a ridge penalized linear regression."}
knitr::include_graphics("images/schema.png/image1.png")
```

@@

The leaking rate alpha define the rate of update of the nodes. The
closer $\alpha$ is to $1$, the higher the reservoir is sensitive to new
inputs (i.e $u(t)$). Therefore, the reservoir state at time $t+1$
denoted $x(t+1)$ depends on the reservoir state at the previous time
(i.e $x(t)$) and the new inputs (i.e $u(t+1)$). Both $W_{in}$ and $W$
are random matrices of size $Nres \times M$ and $Nres \times Nres$
respectively.

$W_{in}$ is a dense matrix generated using a Bernoulli distribution
where each value can be either $-I_{scale}(m)$ or $I_{scale}(m)$ with an
equal probability where $m = 1, …, M$ corresponds to a given feature in
the input layer. The input scaling, denoted $I_{scale}$, is an
hyperparameter coefficient which can be common to all features from the
input layer or specific to each feature $m$. In that case, the more
important the feature is, the greater should be its input scaling. $W$
is a sparse matrix where values are generated from a Gaussian
distribution $\mathcal{N}(0,1)$. Then the matrix $W$ is scaled
according to the spectral radius, an hyperparameter defining the highest
eigen value of $W$.

@@

The parameter $\lambda$ is the ridge penalization which aims to prevent
overfitting. One can also connect the input layer to the output layer in addition to the reservoir nodes. In that case, $X$ is the accumulation of both such that :

@@

Overall, there are four main hyperparameters to be chosen by the user: the leaking rate which defines the memory of the RC, the input scaling which define the relative importance of the features, the spectral radius which define the connections of the neurons inside the reservoir which in turn define the degree of non-linear combination of features and the ridge penalization which controls the degree of overfitting. The choice of hyperparameter often rely on a wrapper approach where the performance of the RC with different combinations of hyperparameters is evaluated on a train set and the best combination is chosen to forecast on the test set.

@@

```{r covidintro, fig.cap="Hospitalizations, IPTCC and positive PCR of Bordeaux University Hospital.", echo = FALSE}
dfOutcome %>%
  tidyr::pivot_longer(cols = c("hosp", "Positive", "Tested")) %>%
  ggplot2::ggplot(mapping = aes(x = date, y = value)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y") +
  theme_minimal() +
  geom_vline(mapping = aes(color = "train-test sets",
                           xintercept = traintest_date)) +
  labs(color = "") +
  theme(legend.position = "bottom")
```

@@

-   Number of nodes (`units`) : it corresponds to the number of nodes inside the reservoir. Usually, the more the better but more nodes increases the computation time.
-   Leaking rate (`lr`) : the leaking rate corresponds to the balance between the new inputs and the previous state. A leaking rate of 1 only consider information from new inputs.
-   Spectral radius (`sr`): the spectral radius is the maximum absolute eigenvalue of the reservoir connectivity matrix. A small spectral radius induces stable dynamics inside the reservoir, a high spectral radius induces chaotic regimen inside the reservoir.
-   Input scaling (`input_scaling`): the input scaling is a gain applied to the input features of the reservoir.
-   Seed (`seed`): because the reservoir connections are set at random, setting the seed is a good approach to ensure   reproducibility. Another approach would be to use several different reservoir but it increases computation time.

@@

Then we can feed the data to the reservoir and see the activation depending on time of the reservoir. To do so, we first prepare the data and transform it to a matrix :

@@

Numerous nodes within the system exhibit a consistent equilibrium state. The challenge arises when the ridge output layer attempts to acquire knowledge from these nodes, as they do not convey meaningful information. This issue can be attributed to the disparate scales of the features. To address this concern, a practical approach involves normalizing the features by dividing each of them by their respective maximum values, thereby scaling them within the range of `0` to `1` by dividing by the maximum of the absolute value.

@@

```{r nodeactivationgood, fig.cap="20 random node acivation over time. Scaled features."}
# feed them to the reservoir
reservoir_state_stand <- predict_seq(node = reservoir,
                                     X = Xstand,
                                     reset = TRUE)
# plot the output
plot(reservoir_state_stand)
```

@@

To perform this task, we need to modify the training and testing process. Leveraging the inherent echo property of the RC, information from preceding time steps is preserved within the reservoir, effectively endowing the RC with a form of memory. Consequently, the final state vector encapsulates insights gathered from all antecedent states. In the context of the sequence-to-vector encoding task, only this ultimate state is employed. This process is executed as follows:

@@

In this paper, we introduce the R package `reservoirnet`, which serves as a versatile tool for implementing reservoir computing based on `ReservoirPy` Python API. It offers flexibility in defining the reservoir architecture, including options for specifying connections between the input layer and the output layer, as well as variations in input scaling as demonstrated on a real-world use case.

@@

Drawing on the robust foundation of the `ReservoirPy` structure, a well-maintained library, this package inherits its reliability and longevity. Nevertheless, our emphasis has been on refining the primary functionalities, which may leave more advanced users desiring deeper access to core functions, such as the generation of the reservoir random connections. This initial version of `reservoirnet` must evolve in tandem with the growing understanding and adoption of reservoir computing within the R community.

>>>>

Reservoir Computing (RC) is a prominent machine learning method, proposed by @jaeger__2001, @maass_real-time_2002 and @lukosevicius_reservoir_2009
that has gained significant attention in recent years for its ability to
efficiently process information generated by dynamical systems. This
innovative approach leverages the dynamics of a high-dimensional
"reservoir" (we define it below) to perform complex computations and solve various tasks based
on the response of this dynamical system to input signals. RC has demonstrated its efficacy in tackling various challenges, encompassing pattern classification and time series forecasting in applications ranging from electrocardiogram analysis to bird calls @trouvain_canary_2021, language processing @hinaut_real-time_2013, power plants, internet traffic, stock prices, and beyond [@tanaka_recent_2019, @lukosevicius_reservoir_2009].

Originally, the RC paradigm was implemented in artificial firing-rate neurons  (“Echo State Networks”, @jaeger__2001) and spiking neurons ("Liquid State Machine”, @maass_real-time_2002) as a recurrent neural network (RNN) where the internal recurrent connections, denoted as the reservoir, are randomly generated and only the output layer (named “read-out”) is trained. The reservoir projects temporal input signals onto a high-dimensional feature space facilitating the learning of non-linear and temporal interactions. Thus, this recurrent layer contains are high-dimensional non-linear recombination of the inputs and past states: it is a “reservoir of computations” from which useful information can be linearly extracted (or “read-out”) to provide the desired outputs. This offers the advantage of decreasing the computing time while consistently maintaining robust performance compared to conventional RNNs [@vlachas_backpropagation_2020]. In parallel, this RC paradigm gained an increasing interest because of its ability to be implemented not only on classical computers. Indeed, the hidden recurrent layer can be kept untrained and thus a wide range of physical media can be used to replace it. Recently, @tanaka_recent_2019 reviewed this prolific field: from FPGA hardware [@penkovsky_efficient_2018], to spin waves using magnetic properties [@nakane_reservoir_2018], skrymions [@prychynenko_magnetic_2018] or optical implementations [@rafayelyan_large-scale_2020]. This provides interesting and potentially more efficient alternative to traditional machine learning.

RC leverages various hyperparameters to introduce prior knowledge about the relationship between input variables and output targets. However, since the connections within the reservoir are randomly initialized, the same set of hyperparameters may exhibit diverse behaviors across different realizations of the reservoir connections. This unpredictability makes it challenging to anticipate the performance of a particular hyperparameter setting, as identical settings may produce varying outcomes when applied to distinct instances of the reservoir. Moreover, selecting the most suitable hyperparameters often requires researchers to experiment with multiple combinations on a training dataset and evaluate their performance on a separate test set. Although this approach can be resource-intensive and time-consuming, it is a compromise that is acceptable considering the rapid simulation capabilities offered by RC.

*XH : Commentaires généraux sur le paragraphe : il faut ne faut pas insister sur les hyperparamètres comme si c’était négatif. Plutôt dire qu’il y a des optimisations et tricks à connaitre, et que c’est intéressant d’avoir une librairie pour ça, et aussi à des fins de reproductibilité. Dire que c’est un trade-off car c’est les a priori nécessaire qu’on rajoute dans l’algo car on gagne du temps sur l’apprentissage. C’est ambigu de parler de stochaticité ici je trouve, car le processus n’est pas stochastic une fois les poids fixés.*

*TF : j'ai modifié, mieux ?*

The utilization of RC poses challenges on multiple fronts. Firstly, its stochastic nature, attributed to the random aspect of the reservoir, introduces unpredictability into the results, even with well-chosen hyperparameters. Secondly, RC relies on numerous hyperparameters to establish connections between neurons. Guidance on determining these hyperparameters is limited, often requiring users to evaluate the performance of different combinations of hyperparameters on a train set before selecting the optimal combination to forecast on the test set^[In this article, we employ the term "train set" to refer to the combined dataset consisting of both the training and validation sets, which are cycled through in a cross-validation manner.]. Thirdly, there is a current absence of implementation in R, rendering the method challenging for users unfamiliar with Python [@trouvain_reservoirpy_2022] or Julia [@martinuzzi_reservoircomputingjl_2022].

*XH : Je pense que ça embrouille là : c’est train/test/validation Les hyperparamètres sont entrainé sur le train set puis évalués sur le validation set. Et ensuite le test set. Par abus de langage on dit souvent « test set » au lieu de « validation set » quand on optimise les paramètres. Moi ça me choque pas, mais dans un papier vaut mieux utiliser les termes adéquats, ou en tous cas toujours utiliser les memes, donc si tu n’as jamais de validation set (enfin le vrai test set), ça peut passer (pas besoin de tout changer). Met juste une note de bas de page pour préciser que en général on devrait l’appeler le validation set. On verra ce que les reviewers disent, pas besoin de perdre du temps maintenant.*

*TF : ok, j'ai rajouté une note de bas de page ci-dessous*

In this paper, we endeavor to delve into these challenges, offering comprehensive guidance to assist new users in maximizing the benefits of RC. Initially, a broad introduction to reservoir computing is presented, accompanied by a tutorial on its application using `reservoirnet`, an R package built upon the `reservoirPy` Python module [@trouvain_reservoirpy_2020, @trouvain_create_2022, @trouvain_reservoirpy_2022]. Subsequently, we investigate the various challenges associated with the use of RC for forecasting COVID-19 hospitalizations. This exploration includes detailed guidance on the modeling strategy, the selection of hyperparameters, and the implementation process.

@@

RC is a machine learning paradigm which is most often implemented as Echo State Networks (ESNs; @jaeger__2001), the firing-rate neuron version. An ESN is described by three matrices of connectivity: an input layer $W_{in}$, a recurrent layer $W$ and an output layer $W_{out}$. At each time step, the input vector $u_t$ is projected into the reservoir which is also combined with reservoir past state $x(t-1)$ through the recurrent connections. The output $y(t)$ is linearly read-out from the reservoir. Input $W_{in}$ and recurrent $W$ matrices are kept random; only the output matrix $W_{out}$ is trained in an offline or online method. Often a ridge regression (i.e. a regularized linear regression) is used to obtain the desired outputs $y(t)$ from the reservoir states $x(t)$. Figure \@ref(fig:rcpresentation) depicts the architecture. For simplicity, we will use the term "reservoir computing” for “Echo State Network” in the remainder of the paper.

```{r rcpresentation, echo=FALSE, fig.cap="my caption", out.width = '90%', fig.cap="Reservoir computing is composed of an input layer, a reservoir and an output layer. Connection between input layer and reservoir and inside reservoir are random. Only the output layer is optimized based on a ridge regression."}
knitr::include_graphics("images/schema.png/image1.png")
```

@@

The leaking rate $\alpha$ defines the rate of update of the nodes. The
closer $\alpha$ is to $1$, the higher the reservoir is sensitive to new
inputs (i.e $u(t)$). Therefore, the reservoir state at time $t+1$
denoted $x(t+1)$ depends on the reservoir state at the previous time
(i.e $x(t)$) and the new inputs (i.e $u(t+1)$). Both $W_{in}$ and $W$
are random matrices of size $N_{res} \times M$ and $N_{res} \times N_{res}$
respectively.

The leaking rate $\alpha$ defines the update rate of the nodes. The closer $\alpha$ is to $1$, the more the reservoir is sensitive to new inputs $u(t)$. Therefore, the reservoir state at time $t+1$ denoted $x(t+1)$ depends on the reservoir state at the previous time $x(t)$ and the new inputs $u(t+1)$. Both $W_{in}$ and $W$ are random matrices of size $N_{res} \times M$ and $N_{res} \times N_{res}$ respectively.

*XH : Ah tu utilises une matrice dense ? si oui, t’as fait ça dans toutes tes simulations ? habituellement on utilise plutôt 10 ou 20% de connectivité, surtout si on prend un distribution bimodale comme bernouilli. En tous cas il faut pas préciser que soit doit être dense ici.*

*TF : on a pris les paramètres par défaut de reservoirpy, si j'ai bien compris, la matrice est générée par la fonction reservoirpy.mat_gen.bernoulli qui par défaut est dense (connectivity = 1)*

$W_{in}$ is a dense matrix generated using a Bernoulli (bimodal) distribution where each value can be either $-I_{scale}(m)$ or $I_{scale}(m)$ with an equal probability where  $m = 1, …, M$ corresponds to a given feature in the input layer. The input scaling, denoted $I_{scale}$, is a hyperparameter coefficient often common to all features from the input layer or specific to each feature $m$. In that case, the more important the feature is, the greater should be its input scaling. $W$ is a matrix (usually sparse) where values are generated from a Gaussian distribution $\mathcal{N}(0,1)$. Then, the $W$ matrix is scaled according to the defined spectral radius, a hyperparameter defining the highest eigen value of $W$.

@@

The parameter $\lambda$ is the ridge penalization which aims to prevent overfitting. Additionally, one can also connect the input layer to the output layer to the reservoir nodes. In that case, $X$ is the accumulation of both such that :

@@

Overall, there are four main hyperparameters to be chosen by the user: the leaking rate which defines the memory of the RC, the input scaling which define the relative importance of the features, the spectral radius which define the connections of the neurons inside the reservoir which in turn define the degree of non-linear combination of features and the ridge penalization which controls the degree of overfitting. The choice of hyperparameter often requires the user to evaluate the performance of different combinations of hyperparameters on a validation set before selecting the optimal combination to forecast on the test set.

@@

```{r covidintro, fig.cap="Hospitalizations, IPTCC and positive PCR of Bordeaux University Hospital.", echo = FALSE}
dfOutcome %>%
  tidyr::pivot_longer(cols = c("hosp", "Positive", "Tested")) %>%
  ggplot2::ggplot(mapping = aes(x = date, y = value)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y") +
  theme_minimal() +
  labs(color = "") +
  theme(legend.position = "bottom")
```

@@

-   Number of nodes (`units`) : it corresponds to the number of nodes inside the reservoir. Usually, the more the better, but more nodes increases the computation time.
-   Leaking rate (`lr`) : the leaking rate corresponds to the balance between the new inputs and the previous state. A leaking rate of 1 only consider information from new inputs.
-   Spectral radius (`sr`): the spectral radius is the maximum absolute eigenvalue of the reservoir connectivity matrix. A small spectral radius induces stable dynamics inside the reservoir, a high spectral radius induces a chaotic regime inside the reservoir.
-   Input scaling (`input_scaling`): the input scaling is a gain applied to the input features of the reservoir.
- Warmup (`warmup`) : it corresponds to the number of time step during which the data are propagating into the reservoir but not used to fit the output layer. This hyperparameter is set in the `reservoirR_fit()` function.

In addition, we can set the seed (`seed`). Because the reservoir connections are set at random, setting the seed is a good approach to ensure reproducibility.

@@

Then we can feed the data to the reservoir and see the activation state of the reservoir $x(t)$. To do so, we first prepare the data and transform it to a matrix:

@@

Numerous nodes within the system exhibit a consistent equilibrium state. The challenge arises when the output layer attempts to extract knowledge from these nodes, as they do not convey meaningful information. This issue can be attributed to the disparate scales of the features. To address this concern, a practical approach involves normalizing the features by dividing each of them by their respective maximum values, thereby scaling them within the range of $-1$ to $1$ by dividing by the maximum of the absolute value. Of note, here the features will be scaled between $0$ and $1$ because all features are positive.

@@

```{r nodeactivationgood, fig.cap="20 random node activation over time. Scaled features."}
# feed them to the reservoir
reservoir_state_stand <- predict_seq(node = reservoir,
                                     X = Xstand,
                                     reset = TRUE)
# plot the output
plot(reservoir_state_stand)
```

@@

To perform this task, we need to modify the training and testing processes. Leveraging the inherent inertia of the reservoir, information from preceding time steps is preserved, effectively endowing the RC with a form of memory. Consequently, the final state vector encapsulates insights gathered from all antecedent states. In the context of the sequence-to-vector encoding task, only this ultimate state is employed. This process is executed as follows:

*XH : Markovian or “echo state” property ? c’est pour dire que les états finissent par être oubliés au bout d’un certain temps, donc c’est pas ce que tu veux dire là je pense*

*TF : j'ai remplacé par inertia qui parait effectivement plus approprié. Mon idée c'est que l'information s'accumule dans le réservoir au fur et à mesure qu'il "lit" la séquence de données.*

@@

In this paper, we introduce the R package `reservoirnet`, which serves as a versatile tool for implementing reservoir computing based on `ReservoirPy`'s Python API. It offers flexibility in defining the reservoir architecture, including options for specifying connections between the input layer and the output layer, as well as variations in input scaling as demonstrated on a real-world use case.

@@

Drawing on the robust foundation of the `ReservoirPy` structure, a well-maintained library, this package inherits its reliability and longevity. We have focused on providing access to the fundamental features, building upon the strong base provided by `ReservoirPy.` Therefore, this initial version of `reservoirnet` must evolve in tandem with the growing understanding and adoption of reservoir computing within the R community.

*XH : Il faut rephrase mais je suis pas sur de ce qui tu veux dire.*

*TF : j'ai reformulé ce dernier paragraphe en espérant que ça soit mieux. Mon idée c'était d'expliquer qu'on a focalisé le package sur les fonctions principales mais que tout n'est pas dispo depuis R (par exemple on ne peut pas choisir comment sont générées les matrices de connection aléatoires du réservoir, on les a laissé aux paramètres par défaut).*
